An Intrinsically-Typed Solution for the\\ List-Machine Benchmark

### 2021-03-10 papers/COLA/paper.tex
More updates

\subsection{Modification in the intepreter}\label{sec:changesintep}

@@

>>>>

Most of the code for the type-checker is ummodified in our new version. We just need to change
the type checking for jumps, which now should validate if a machine register has an appropriate
continuation type.

We also need to include code to check the well-typedness for the get-label instruction. First,
we include a new constructor on type \ensuremath{\D{UInstr}} to represent untyped get-label instruction and
add its corresponding equation on function \ensuremath{\F{forge\textrm{-}types\textrm{-}instr}}, which erases type information
from the instrinsically typed syntax. This function is used by the type-checker to ensure its
correctness.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{17}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\mathkw{data}\;\D{UInstr}\;\mathbin{:}\;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\orange{\texttt{-- same code from before.}}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\Con{get\textrm{-}label}\;{}\<[17]%
\>[17]{}\mathbin{:}\;(\V{v}\;\mathbin{:}\;\V{ℕ})\;\rightarrow\;\D{Label}\;\rightarrow\;\D{UInstr}{}\<[E]%
\\[\blanklineskip]%
\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{\Pi}\;\V{\Gamma}\;\V{\Gamma'}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{\vdash}\;\V{\Gamma}\;\D{\Rightarrow}\;\V{\Gamma'}\;\rightarrow\;\D{UInstr}{}\<[E]%
\\
\>[B]{}\orange{\texttt{-- same code as before...}}{}\<[E]%
\\
\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel\textrm{-}0}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\Con{L_{0}}{}\<[E]%
\\
\>[B]{}\F{forge\textrm{-}types\textrm{-}instr}\;(\Con{instr\textrm{-}getlabel}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\{\mskip1.5mu \V{x}\;\mathrel{=}\;\V{x}\mskip1.5mu\}\;\V{v}\;\V{p})\;\mathrel{=}\;\Con{get\textrm{-}label}\;\V{x}\;\V{l}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
Constructor \ensuremath{\Con{get\textrm{-}label}} stores the name for a register and a label which is loaded into the register.
Equations for erasing type information are immediate. The only peculiarity is that when the instruction
\ensuremath{\Con{get\textrm{-}label}} is typed using rule \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}} we consider the initial label \ensuremath{\Con{L_{0}}}. Next, we implement
the type checker logic for \ensuremath{\Con{get\textrm{-}label}}. Function \ensuremath{\F{check\textrm{-}get\textrm{-}label}} verifies if the label is present at
program global typing context and if the variable is present on current block typing context.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{30}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\mathbin{:}\;\D{\forall}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;{}\<[30]%
\>[30]{}\rightarrow\;\D{TC}\;(\D{CheckedInstr}\;\V{\Pi}\;\V{\Gamma}\;(\Con{get\textrm{-}label}\;\V{v}\;\V{l})){}\<[E]%
\\
\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;\mathkw{with}\;\F{inspect}\;(\F{lookup}\;\V{\Pi}\;\V{l}){}\<[E]%
\\
\>[B]{}\F{check\textrm{-}get\textrm{-}label}\;\V{\Pi}\;\V{\Gamma}\;\V{v}\;\V{l}\;\mid \;\V{\Gamma_1}\;\F{with\equiv}\;\V{p}\;\mathkw{with}\;\F{lookup\textrm{-}var}\;\V{\Gamma}\;\V{v}{}\<[E]%
\\
\>[B]{}\V{...}\;\mid \;\Con{nothing}\;\mathrel{=}\;\F{type\textrm{-}error}\;\text{\ttfamily \char34 check-get-label:~variable~out~of~scope\char34}{}\<[E]%
\\
\>[B]{}\V{...}\;\mid \;\Con{just}\;(\V{\tau}\;\red{,}\,\;\V{idx})\;\mathrel{=}\;\Con{right}\;(\Con{ok}\;(\Con{instr\textrm{-}getlabel}\;\V{idx}\;(\F{lookup⇒[]=}\;\V{l}\;\V{\Pi}\;\V{p}))){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

\subsection{Modifications in the intepreter}\label{sec:changesintep}

The final piece of our formalization for the version 2.0 of the list machine benchmark is modifying
the interpreter implementation. Our first step is to change the definition of values to reflect
the changes on the type syntax, since now we have a new value for continuation types. A continuation
value simply stores the label of the corresponding block.

\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[3]{}\mathkw{data}\;\D{Val}\;\mathbin{:}\;\D{Ty}\;\rightarrow\;\D{Set}\;\mathkw{where}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\orange{\texttt{-- same code as before}}{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\Con{cont}\;\mathbin{:}\;\D{\forall}\;\{\mskip1.5mu \V{l}\;\V{\Gamma_1}\mskip1.5mu\}\;\rightarrow\;\V{\Pi}\;\D{[}\;\V{l}\;\D{]=}\;\V{\Gamma_1}\;\rightarrow\;\D{Val}\;(\Con{cont}\;\V{\Gamma_1}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

Since the value definition has changed, we need to modify the lemmas for value and environment
subsumption accordingly. Next, we change the interpreter itself by adding equations for the
\ensuremath{\Con{get\textrm{-}label}} instruction and fixing the \ensuremath{\Con{jump}} implementation.
Equations for \ensuremath{\Con{get\textrm{-}label}} simply update the current environment using the label. When the
typed syntax is built using constructor \ensuremath{\Con{instr\textrm{-}getlabel\textrm{-}0}}, the environment is updated with
value \ensuremath{\Con{nil}}, which denotes the initial continuation for the program execution.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{penv}\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}getlabel\textrm{-}0}\;\V{idx})\;\V{b})\;{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{penv}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{idx}\;\Con{nil})\;\V{b}{}\<[E]%
\\
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{penv}\;\V{p}\;\V{env}\;(\Con{block\textrm{-}seq}\;(\Con{instr\textrm{-}getlabel}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{idx}\;\V{x})\;\V{b})\;{}\<[E]%
\\
\>[B]{}\hsindent{5}{}\<[5]%
\>[5]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{penv}\;\V{p}\;(\F{update\textrm{-}env}\;\V{env}\;\V{idx}\;(\Con{cont}\;\V{x}))\;\V{b}{}\<[E]%
\ColumnHook
\end{hscode}\resethooks
The final piece of changes in the interpreter is implementing jump instruction. In order to
execute an indirect jump, first we need to get the continuation value associated with its
variable using \ensuremath{\F{lookup}\;\V{env}\;\V{v}}. Using the label associated with the retrieved continuation value
we jump to its corresponding block and its entry execution environment.
\begin{hscode}\SaveRestoreHook
\column{B}{@{}>{\hspre}l<{\hspost}@{}}%
\column{3}{@{}>{\hspre}l<{\hspost}@{}}%
\column{5}{@{}>{\hspre}l<{\hspost}@{}}%
\column{E}{@{}>{\hspre}l<{\hspost}@{}}%
\>[B]{}\F{run\textrm{-}step}\;(\Con{suc}\;\V{n})\;\V{penv}\;\V{p}\;\V{env}\;(\Con{block\textrm{-}jump}\;\V{v}\;\V{x₁})\;\mathkw{with}\;\F{lookup}\;\V{env}\;\V{v}{}\<[E]%
\\
\>[B]{}\hsindent{3}{}\<[3]%
\>[3]{}\V{...|}\;(\Con{cont}\;\{\mskip1.5mu \V{l}\;\mathrel{=}\;\V{l}\mskip1.5mu\}\;\V{k})\;\mathkw{rewrite}\;\D{sym}\;(\F{[]\textrm{=}{\Rightarrow}lookup}\;\V{k})\;{}\<[E]%
\\
\>[3]{}\hsindent{2}{}\<[5]%
\>[5]{}\mathrel{=}\;\F{run\textrm{-}step}\;\V{n}\;\V{penv}\;\V{p}\;(\F{lookupA}\;\V{l}\;\V{penv})\;(\F{lookupA}\;\V{l}\;\V{p}){}\<[E]%
\ColumnHook
\end{hscode}\resethooks

@@

Appel's list machine benchmark 2.0 uses a semantic-based approach to prove type soundness and the
correctness of the type checker~\cite{AppelMRV07}, which rely on a Coq library inspired by modal logics. We stick to our initial
proposal based on dependently typed syntax and definitional interpreters.
